---
title: "The R API for the mzTab-M File Format"
package: RmzTabM
format:
  html:
    minimal: true
    theme: flatly
vignette: >
  %\VignetteIndexEntry{The R API for the mzTab-M File Forat}
  %\VignetteKeywords{Mass Spectrometry, MS, MSMS, Metabolomics, Infrastructure, Quantitative}
  %\VignettePackage{RmzTabM}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
  %\VignetteDepends{RmzTabM,BiocStyle}

---

# Introduction

# Installation

# R mzTab-M API

... TODO ... general info, links to information ...

```{r}
library(RmzTabM)
```

## High-level, user faced functions

## Low-level functions

The low-level functions listed in this section provide the functionality to
convert or format information and data for/from the [mzTab-M] format. These
functions are designed to be re-used by other R packages and should take and
return only basic, plain R data types.

### Formatting and exporting

#### Metadata

The mzTab-M format defines various fields and parameters to describe the data
and information of an experiment. The *RmzTabM* package provides a variety of
utility functions that help defining and formatting this information.

### Reading and importing


# Session information

```{r}
sessionInfo()
```
