---
title: "The R API for the mzTab-M File Format"
package: RmzTabM
format:
  html:
    minimal: true
    theme: flatly
vignette: >
  %\VignetteIndexEntry{The R API for the mzTab-M File Format}
  %\VignetteKeywords{Mass Spectrometry, MS, MSMS, Metabolomics, Infrastructure, Quantitative}
  %\VignettePackage{RmzTabM}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
  %\VignetteDepends{RmzTabM,BiocStyle,pander}
---

# Introduction

The *RmzTabM* package provides the API and core functionality to read and write
files in mzTab-M format. The functions can be re-used and integrated by other R
packages to support import and export of their respective
metabolomics/lipidomics result objects in this format.

# Installation

# R mzTab-M API

The *RmzTabM* package provides low level, core functions and higher-level
functions to work with files in mzTab-M format. The high-level functions are
more user-oriented, simplifying the import and export of data and information
from and to files in mzTab-M format. The low-level, core functions are
developer-oriented, providing helper functions that can be re-used in other R
packages to read and write from and to mzTab-M files.

For a description of the mzTab-M format and the set of mandatory and optional
fields refer to the official [format
definition](https://github.com/HUPO-PSI/mzTab-M/blob/main/specification_documents/mzTab_format_specification_2_1-M.adoc).

```{r}
library(RmzTabM)
```

## High-level, user faced functions

TODO: implement these functions. These functions should simplify import/export
taking more complex or multiple data parts (`data.frame`s, `matrix` etc) as
input and write the formatted data directly to a mzTab-M file, or should read a
mzTab-M file returning it's content as e.g. a `list` of elements.

## Low-level functions

The low-level functions listed in this section provide the base functionality to
convert or format information and data for/from the [mzTab-M
format](https://github.com/HUPO-PSI/mzTab-M/blob/main/specification_documents/mzTab_format_specification_2_1-M.adoc). These
functions are designed to be re-used by other R packages and take and return
only basic, plain R data types.

### Formatting and exporting

#### Metadata

The mzTab-M format defines various fields and parameters to describe the data
and information of an experiment. The *RmzTabM* package provides a variety of
utility functions that help defining and formatting this information.

See also the [specification of the MTD
section](https://github.com/HUPO-PSI/mzTab-M/blob/main/specification_documents/mzTab_format_specification_2_1-M.adoc#62-metadata-section)
for more information and optional and mandatory metadata fields.

The general categories of the metadadata in the mzTab-M MTD section are *core
information*, *sample information*, *MS run information*, *assay information*
and *study variable information*. For each of these categories a separate R
function is available to create and format the respective fields. As an example,
we define below a `data.frame` with sample information. In our example we assume
3 samples (e.g. cell lines) each measured at two different time points. An
additional column *genotype* specifies the genotype of the individual samples
and a column *operator* the initials of the researcher extracting the samples.

```{r}
#' Define a simple data.frame of the measured samples of an experiment
exp <- data.frame(
    sample_name = c("S1_T1", "S1_T2", "S2_T1", "S2_T2", "S3_T1", "S3_T2"),
    sample_id = c("S1", "S1", "S2", "S2", "S3", "S3"),
    timepoint = c("0h", "6h", "0h", "6h", "0h", "6h"),
    genotype = c("WT", "WT", "KO", "KO", "KO", "KO"),
    operator = c("BB", "BB", "BB", "BB", "FB", "FB"),
    file_name = c("s1-t1.mzML", "s1-t2.mzML", "s2-t1.mzML", "s2-t2.mzML",
                  "s3-t1.mzML", "s3-t2.mzML")
)
exp
```

We will next compile the MTD information for the experiment using the individual
helper functions, starting with the **Core information**: this comprises general
information about the experiment. A minimal set of fields can be compiled using
the `mtd_skeleton()` function. We have to provide an ID for the experiment and
in addition we specify the software used to process the data:

```{r}
mtd <- mtd_skeleton(
    id = "EXP_001",
    software = "[MS, MS:1001582, xcms, 4.1.0]"
)
mtd
```

This represents some minimal information. The data of the MTD section is
formatted as a `character` 2-column `matrix`. We could now either change the
value (i.e., the elements in the second column of this matrix) of existing
fields, or also manually add additional fields/information. As an example we add
a title and description for the experiment. See also the
[mzTab-M](https://github.com/HUPO-PSI/mzTab-M/blob/main/specification_documents/mzTab_format_specification_2_1-M.adoc#62-metadata-section)
format definition for other supported fields.

```{r}
mtd <- rbind(
    mtd,
    c("title", "Experiment 1 preprocessed data"),
    c("description", "The preprocessed data of the experiment 1.")
)
```

To help with formatting we can also use the `mtd_fields()` function. Below we
use this function to add information about the MS instrumentation to the MTD
section:

```{r}
instr <- mtd_fields(
    name = "[MS, MS:1000449, LTQ Orbitrap,]",
    source = "[MS, MS:1000073, ESI,]",
    `analyzer[1]` = "[MS, MS:1000291, linear ion trap,]",
    detector = "[MS, MS:1000253, electron multiplier,]",
    field_prefix = "instrument"
)
instr
```

And we add that information to the `mtd` variable.

```{r}
mtd <- rbind(mtd, instr)
```

Note that the general information part should also contain the references to
**all** controlled vocabulary (CV) ontologies used in the mzTab-M file. The
default ontologies added by the `mtb_skeleton()` function are the *PSI-MS* and
the *PRIDE* ontologies. If other vocabularies are used, they should be added
manually (following the scheme of the others, i.e., the fields starting with
`"cv["`).

The next category of metadata information is **sample information**. This
comprises (optional) information on individual samples that were measured with
the various assays/runs. We use the `mtd_sample()` function to assist in
compiling this information. Parameters `sample`, `species`, `tissue` and
`cell_type`, `disease` and `description` allow to provide pre-defined sample
properties. Additional sample annotations and details can be provided through
the function's `...`. For the example below we define some of these properties
and in addition provide a custom field for the extraction data. Be aware that
`mtd_sample()` does **not** support partial or positional matching of
parameters; for each of the parameters the **full parameter name** has to be
used (i.e., `sample = ...` instead of `sam = ...` or `s = ...`).

```{r}
mtd_s <- mtd_sample(
    sample = unique(exp$sample_id),
    species = "[NCBITaxon, NCBITaxon:9606, Homo sapiens, ]",
    tissue = "[BTO, BTO:0000759, liver, ]",
    cell_type = "[CL, CL:0000182, hepatocyte, ]",
    c("[,,Extraction date, 2011-12-21]",
      "[,,Extraction date, 2011-12-22]",
      "[,,Extraction date, 2011-12-23]")
    )
mtd_s
```

We can then add this information to the `mtd` variable by simply `rbind()`ing
it.

```{r}
mtd <- rbind(mtd, mtd_s)
```

Next we compile **MS run information** of the experiment using the
`mtd_ms_run()` helper function. This should comprise all (MS-specific)
information related to the measurement of each sample - including also the MS
data file names and locations. For our example we use the file names
reported in the sample data frame and specify the polarity of the measurement
runs.

```{r}
mtd_msr <- mtd_ms_run(
    location = exp$file_name,
    format = "[MS, MS:1000584, mzML file, ]",
    id_format = "[MS, MS:1000530, mzML unique identifier, ]",
    scan_polarity = "positive")
mtd_msr
```

Each row in the `exp` data frame was assigned to a `"ms_run"` with the location
and format of the respective file as well as the polarity in which the data was
acquired. We can combine this data with the `mtd` variable.

```{r}
mtd <- rbind(mtd, mtd_msr)
```

Next we define the **assay information**. Generally, each measurement (MS run)
is associated to one assay, but also more complex configurations are
supported. See the help of the `mtd_assay()` function for details on multiplexed
or pre-fractionated samples. Mandatory information that has to be provided to
the `mtd_assay()` function are the name (ID) of the assay and the reference to
the MS run in which the assay was measured. For the latter, a format of
`"ms_run[<index of the MS run>]"` is expected. For our example we provide in
addition also the (optional, but suggested) reference to the original
sample. Note that each assay must represent one column in the following feature
abundance table (SMF).

```{r}
mtd_a <- mtd_assay(
    assay = exp$sample_name,
    sample_ref = c("sample[1]", "sample[1]", "sample[2]", "sample[2]",
                   "sample[3]", "sample[3]"),
    ms_run_ref = paste0("ms_run[", seq_len(nrow(exp)), "]")
)
mtd_a
```

We add this information to the `mtd` variable.

```{r}
mtd <- rbind(mtd, mtd_a)
```

At last we compile the **study variable information** of our example
experiment. This should capture all experiment-relevant study variables. In R,
such information is generally encoded in a *sample* or *phenotype* `data.frame`,
with rows being individual samples (or measurements thereof) and columns the
sample characteristics (i.e., the study variables). For our example we use the
`mtd_study_variables()` function to generate the mzTab-M study variable
information, specifying with parameter `study_variable_columns` the names of the
`data.frame` containing the relevant information.

```{r}
mtd_svar <- mtd_study_variables(
    exp,
    study_variable_columns = c("timepoint", "genotype", "operator"))
mtd_svar
```

Each unique value in each of the specified columns is encoded as a
`study_variable` (or rather as a *study variable value*), with its *assay_refs*
attribute containing the rows/assays in which this value was measured. The
variable's *description* (by default) indicates the name of the column. The
*average_function* and *variation_function* attributes allow to define the
function that was used to calculate the average and variance of the abundance
values for that variable value. In upcoming versions a dedicated attribute
*group* can be used to specify the name of the column (i.e. the study variable
name) the study variable value is provided in.

We next add the study variable information to the `mtd`variable.

```{r}
mtd <- rbind(mtd, mtd_svar)
```

At last we sort the elements according to the expected order in the MTD
section using the `mtd_sort()` function.

```{r}
mtd <- mtd_sort(mtd)
```

This two-column `matrix` could now be saved to a text file using a tabulator
(`"\t"`) as a field separator. The full metadata header is shown in the table
below.

```{r, results = "asis"}
library(pander)
pandoc.table(mtd, style = "rmarkdown", split.table = Inf)
```


### Reading and importing

TODO: implement these functions moving the respective code from the legacy repo

### General utility functions

General utility functions include:

- `mtd_fields()`: to format values in the mzTab-M-specific format.
- `mtd_sort()`: to sort rows of the metadata `matrix` into the expected order.

# Session information

```{r}
sessionInfo()
```
