---
title: "mzTab-M import/export playground"
---

Code examples to export and import mzTab-M files using *RmzTabM*. This is the
playground to test implementations for higher-level import/export functionality
and to evaluate that results are compatible with the standard.

# Exporting through *RmzTabM*

## *xcms* result objects

### FAAH KO example file

Loading an example *xcms* result object.

```{r}
library(MsExperiment)
library(xcms)
xmse <- loadXcmsData()

#' sample/metadata information:
#'
#' Add additional column for "experiment", but then restrict to:
#' - sample_group
#' - experiment
#' - weight
#' - sample_name (? really?)
sampleData(xmse)$experiment <- c("exp1", "exp2", "exp3", "exp4",
                                 "exp1", "exp2", "exp3", "exp4")
sampleData(xmse)$weight <- c(45.2, 56.3, 36.1, 55.1, 62.4, 65.2, 58.2, 56.8)
sampleData(xmse)

#' feature (SMF) information:
#'
#' use mz and rt information, don't consider the others
featureDefinitions(xmse)

#' abundance values:
featureValues(xmse, method = "sum") |> head()

```

We next export this data to an mzTab-M file using *RmzTabM*.
In order todo:
    - Create MTD table.
    - Create SML table. (all `"null"` since is not exported by *xcms*)
    - Create SMF table.
    - Create SME table. (For the moment do not export. Prepare the check)
    - Create a unique df with all section.
    - Write tsv


#### Create MTD table
```{r}
devtools::load_all()

## Parameter
id_exp <- "test_xcms_save"
xmse <- xmse
title <- character()
description <- character()
instrument = NULL
col_species = character()
col_tissue = character()
col_cell_type = character()
col_disease = character()
col_description = character()
ms_run_polarity = character()
groups = c("sample_group", "experiment", "sample_type")
group_unit = c("", "", "")

## TODO?: add all parameters of sub functions?
mtd_xcms_export <- function(id_exp, xmse, title = character(),
            description = character(), instrument = NULL,
            col_species = character(), col_tissue = character(),
            col_cell_type = character(), col_disease = character(),
            col_description = character(), ms_run_polarity = character(),
            groups = character(), group_unit = character(),
            group_description = character(), group_type = character(),
            group_datatype = character(),
            average_function = "[MS, MS:1002962, mean, ]",
            variation_function = "[MS, MS:1002963, variation coefficient, ]",
            svar_description = character()){
    ## MTD section: sampleData()
    mtd_sample_data <- sampleData(xmse)

    ## TODO: expand with the other fields.
    xcms_v <- paste0("[MS, MS:1001582, xcms, ", packageVersion("xcms"), "]")
    mtd <- mtd_skeleton(
        id = id_exp,
        software = xcms_v
    )
    if(length(title))
        mtd <- rbind(mtd, c("title", title))
    if(length(description))
        mtd <- rbind(mtd, c("description", description))

    ## INSTRUMENT
    if(!is.null(instrument)) {
        instr <- .mtd_get_field(instrument, "^instrument\\[\\d+\\]",
                                exact = FALSE, fixed = FALSE)[[1]]
        if (any(is.na(instr)))
            wanirng("No valid \"instrument\" fields detected in the
                    \"instrument\" parameter. Skip the saving.")
        else {
            rbind(mtd, instrument[instrument[, 1] %in% names(instr), ])
        }
    }

    ## SAMPLE INFORMATION
    species = ifelse(length(col_species), mtd_sample_data[, eval(col_species)],
                     list())
    tissue = ifelse(length(col_tissue), mtd_sample_data[, eval(col_tissue)],
                     list())
    cell_type = ifelse(length(col_cell_type),
                        mtd_sample_data[, eval(col_cell_type)],
                        list())
    disease = ifelse(length(col_disease), mtd_sample_data[, eval(col_disease)],
                     list())
    description_sample = if(length(col_description)){
                            mtd_sample_data[, eval(col_description)]
                        } else {character()}
    mtd_s <- mtd_sample(sample = mtd_sample_data$sample_name,
                        species = species,
                        tissue = tissue,
                        cell_type = cell_type,
                        disease = disease,
                        description = description_sample)
    mtd <- rbind(mtd, mtd_s)
    ## Add sample[1-n] to correct sample in mtd_sample_data
    sample_ref <- .mtd_get_field(mtd_s, "^sample\\[\\d+\\]$",
                                 exact = FALSE, fixed = FALSE)[[1]]
    mtd_sample_data$sample_ref <- names(sample_ref[match(sample_ref,
                                                mtd_sample_data$sample_name)])

    ## MS_RUNS
    ## TODO: get correct scan_polarity from xcms
    ## TODO: parse all parameters from xcms
    mtd_msr <- mtd_ms_run(location = mtd_sample_data$spectraOrigin,
                            scan_polarity = ms_run_polarity)
    mtd <- rbind(mtd, mtd_msr)
    ## Add ms_runs[1-n] to correct sample in mtd_sample_data
    ms_runs_ref <- .mtd_get_field(mtd_msr, "^ms_run\\[\\d+\\]-location$",
                                exact = FALSE, fixed = FALSE)[[1]]
    mtd_sample_data$ms_runs_ref <- names(ms_runs_ref[match(ms_runs_ref,
                                                        mtd_sample_data$spectraOrigin)])
    mtd_sample_data$ms_runs_ref <- unlist(lapply(strsplit(mtd_sample_data$ms_runs_ref,
                                                    "-"), function(x){x[1]}))

    ## ASSAY
    ## sample_ref not mandatory
    mtd_a <- mtd_assay(assay = mtd_sample_data$sample_name,
                        sample_ref = mtd_sample_data$sample_ref,
                        ms_run_ref = mtd_sample_data$ms_runs_ref)
    mtd <- rbind(mtd, mtd_a)

    ## STUDY VARIABLES
    mtd_svar <- mtd_study_variables(mtd_sample_data,
                                    groups = groups,
                                    group_unit = group_unit,
                                    group_description = group_description,
                                    group_type = group_type,
                                    group_datatype = group_datatype,
                                    average_function = average_function,
                                    variation_function = variation_function,
                                    description = svar_description)
    mtd <- rbind(mtd, mtd_svar)

    mtd <- mtd_sort(mtd)
    mtd
}
```

#### Create SML

```{r}
sml_xcms_export <- function(xmse, mtd) {
    ## To implement since right now is not exported byt xcms
    abundance <- matrix("null", nrow = 1, ncol = nrow(sampleData(xmse)))
    sml <- sml_create(x = abundance)
    ## TODO: put a random reliability to pass validator.
    ## TODO: create issue in the valdiator
    sml$reliability <- 4
    sml <- sml_add_study_variable_columns(sml, mtd)
    sml[is.na(sml)] <- "null"
    sml
}

```

#### Create SMF

```{r}
matrix_integer_to_character <- function(m) {
    if (!is.matrix(m)) stop("Input must be a matrix")
    out <- matrix(as.character(m), nrow = nrow(m), ncol = ncol(m),
                    dimnames = dimnames(m))
    out[is.na(m)] <- "null"
    out
}

smf_xcms_export <- function(xmse) {
    ## use mz and rt information, don't consider the others
    feature_info <- featureDefinitions(xmse)
    ## abundance values:
    abundance <- matrix_integer_to_character(featureValues(xmse,
                                                           method = "sum"))

    charge = ifelse("charge" %in% names(feature_info), feature_info$charge,
                    character())
    adduct_ion = ifelse("adduct_ion" %in% names(feature_info),
                        feature_info$adduct_ion, character())
    isotopomer = ifelse("isotopomer" %in% names(feature_info),
                        feature_info$isotopomer, character())
    ## TODO: add SME_ID_REFS, SME_ID_REF_ambiguity_code
    smf <- smf_create(x = abundance,
                        exp_mass_to_charge = feature_info$mzmed,
                        retention_time_in_seconds = feature_info$rtmed,
                        retention_time_in_seconds_start = feature_info$rtmin,
                        retention_time_in_seconds_end = feature_info$rtmax,
                        charge = charge,
                        adduct_ion = adduct_ion,
                        isotopomer = isotopomer,
                        global_feature_id = rownames(feature_info))
    smf
}
```

#### Create SME

```{r}
sme_xcms_export <- function(xmse) {
    ## TODO: add SME. Default empty data.frame
    sme <- data.frame()
    sme
}
```


##### Create unique table and save tsv

```{r}
mtd <- mtd_xcms_export("xcms_mzTab-M_2-1", xmse,
                        ms_run_polarity = "positive",
                        groups = c("sample_group", "experiment"),
                        group_unit = c("", ""))
sml <- sml_xcms_export(xmse, mtd)
smf <- smf_xcms_export(xmse)
sme <- sme_xcms_export(xmse)

path <- "./inst/mztabm/out"
if(!dir.exists(path))
    dir.create(path, showWarnings = FALSE)

filename = file.path(path,
                    paste0(.mtd_get_field(mtd, "^mzTab-ID$",
                                        exact = FALSE, fixed = FALSE),
                    ".mzTab"))

write.table(cbind(data.frame(rep("MTD", nrow(mtd))), mtd),
            file = filename, sep = "\t", row.names = FALSE,
            col.names = FALSE, quote = FALSE, fileEncoding = "UTF-8")
write.table("", file = filename, sep = "\t", row.names = FALSE,
            col.names = FALSE, append = TRUE, quote = FALSE,
            fileEncoding = "UTF-8")
suppressWarnings(write.table(sml, file = filename, sep = "\t",
                             row.names = FALSE, append = TRUE, quote = FALSE,
                             fileEncoding = "UTF-8"))
write.table("", file = filename, sep = "\t", row.names = FALSE,
            col.names = FALSE, append = TRUE, quote = FALSE,
            fileEncoding = "UTF-8")
if(nrow(smf)) {
    suppressWarnings(write.table(smf, file = filename, sep = "\t",
                                 row.names = FALSE, append = TRUE,
                                 quote = FALSE, fileEncoding = "UTF-8"))
    write.table("", file = filename, sep = "\t", row.names = FALSE,
                col.names = FALSE, append = TRUE, quote = FALSE,
                fileEncoding = "UTF-8")
}
if(nrow(sme)) {
    suppressWarnings(write.table(sme, file = filename, sep = "\t",
                                 row.names = FALSE, append = TRUE,
                                 quote = FALSE, fileEncoding = "UTF-8"))
}
```

### *Metabonaut* example file?

Here we would have some tentative annotations (only very few though) and a data
set with more phenotypic information (age, sex). Plus: we could add a dedicated
vignette with the mzTab-M export to *Metabonaut*.

# Importing through *RmzTabM*

# Session information

```{r}
sessionInfo()
```
