---
title: "mzTab-M import/export playground"
---

Code examples to export and import mzTab-M files using *RmzTabM*. This is the
playground to test implementations for higher-level import/export functionality
and to evaluate that results are compatible with the standard.

# Exporting through *RmzTabM*

## *xcms* result objects

### FAAH KO example file

Loading an example *xcms* result object.

```{r}
library(MsExperiment)
library(xcms)
xmse <- loadXcmsData()

#' sample/metadata information:
#'
#' Add additional column for "experiment", but then restrict to:
#' - sample_group
#' - experiment
#' - weight
#' - sample_name (? really?)
sampleData(xmse)$experiment <- c("exp1", "exp2", "exp3", "exp4",
                                 "exp1", "exp2", "exp3", "exp4")
sampleData(xmse)$weight <- c(45.2, 56.3, 36.1, 55.1, 62.4, 65.2, 58.2, 56.8)
sampleData(xmse)

#' feature (SMF) information:
#'
#' use mz and rt information, don't consider the others
featureDefinitions(xmse)

#' abundance values:
featureValues(xmse, method = "sum") |> head()

```

We next export this data to an mzTab-M file using *RmzTabM*...


### *Metabonaut* example file?

Here we would have some tentative annotations (only very few though) and a data
set with more phenotypic information (age, sex). Plus: we could add a dedicated
vignette with the mzTab-M export to *Metabonaut*.

# Importing through *RmzTabM*

# Session information

```{r}
sessionInfo()
```
