---
title: "mzTab-M import/export playground"
---

Code examples to export and import mzTab-M files using *RmzTabM*. This is the
playground to test implementations for higher-level import/export functionality
and to evaluate that results are compatible with the standard.

# Exporting through *RmzTabM*

## *xcms* result objects

### FAAH KO example file

Loading an example *xcms* result object.

```{r}
library(MsExperiment)
library(xcms)
xmse <- loadXcmsData()

#' sample/metadata information:
#'
#' Add additional column for "experiment", but then restrict to:
#' - sample_group
#' - experiment
#' - weight
#' - sample_name (? really?)
sampleData(xmse)$experiment <- c("exp1", "exp2", "exp3", "exp4",
                                 "exp1", "exp2", "exp3", "exp4")
sampleData(xmse)$weight <- c(45.2, 56.3, 36.1, 55.1, 62.4, 65.2, 58.2, 56.8)
sampleData(xmse)

#' feature (SMF) information:
#'
#' use mz and rt information, don't consider the others
featureDefinitions(xmse)

#' abundance values:
featureValues(xmse, method = "sum") |> head()

```

We next export this data to an mzTab-M file using *RmzTabM*.
In order todo:
    - Create MTD table.
    - Create SML table. (all `"null"` since is not exported by *xcms*)
    - Create SMF table.
    - Create SME table. (For the moment do not export. Prepare the check)
    - Create a unique df with all section.
    - Write tsv


#### Create MTD table
```{r}
devtools::load_all()

## TODO:
## - mtd_skeleton() parameter.
## - add all possible mtd_fields(), like the instrument test.
## - mtd_ms_run() parameter.
## - mtd_assay() parameter.
## - mtd_study_variables() parameter. Not all passed as parameters.

## TODO?: add all parameters of sub functions?
## Already too many parameters. Need to manage them better.
mtd_xcms_export <- function(id_exp, xmse, title = character(),
            description = character(), instrument = NULL,
            col_species = character(), col_tissue = character(),
            col_cell_type = character(), col_disease = character(),
            col_description = character(), ms_run_polarity = character(),
            groups = character(), group_unit = character(),
            group_description = character(), group_type = character(),
            group_datatype = character(),
            average_function = "[MS, MS:1002962, mean, ]",
            variation_function = "[MS, MS:1002963, variation coefficient, ]",
            svar_description = character()){
    ## MTD section: sampleData()
    mtd_sample_data <- sampleData(xmse)

    ## TODO: expand with the other fields of mtd_skeleton.
    xcms_v <- paste0("[MS, MS:1001582, xcms, ", packageVersion("xcms"), "]")
    mtd <- mtd_skeleton(
        id = id_exp,
        software = xcms_v
    )
    if(length(title))
        mtd <- rbind(mtd, c("title", title))
    if(length(description))
        mtd <- rbind(mtd, c("description", description))

    ## INSTRUMENT (add other fields)
    if(!is.null(instrument)) {
        instr <- .mtd_get_field(instrument, "^instrument\\[\\d+\\]",
                                exact = FALSE, fixed = FALSE)[[1]]
        if (any(is.na(instr)))
            wanirng("No valid \"instrument\" fields detected in the
                    \"instrument\" parameter. Skip the saving.")
        else {
            rbind(mtd, instrument[instrument[, 1] %in% names(instr), ])
        }
    }

    ## SAMPLE INFORMATION
    species = ifelse(length(col_species), mtd_sample_data[, eval(col_species)],
                     list())
    tissue = ifelse(length(col_tissue), mtd_sample_data[, eval(col_tissue)],
                     list())
    cell_type = ifelse(length(col_cell_type),
                        mtd_sample_data[, eval(col_cell_type)],
                        list())
    disease = ifelse(length(col_disease), mtd_sample_data[, eval(col_disease)],
                     list())
    description_sample = if(length(col_description)){
                            mtd_sample_data[, eval(col_description)]
                        } else {character()}
    mtd_s <- mtd_sample(sample = mtd_sample_data$sample_name,
                        species = species,
                        tissue = tissue,
                        cell_type = cell_type,
                        disease = disease,
                        description = description_sample)
    mtd <- rbind(mtd, mtd_s)
    ## Add sample[1-n] to correct sample in mtd_sample_data
    sample_ref <- .mtd_get_field(mtd_s, "^sample\\[\\d+\\]$",
                                 exact = FALSE, fixed = FALSE)[[1]]
    mtd_sample_data$sample_ref <- names(sample_ref[match(sample_ref,
                                                mtd_sample_data$sample_name)])

    ## MS_RUNS
    ## TODO: get correct scan_polarity from xcms
    ## TODO: parse all parameters from xcms
    mtd_msr <- mtd_ms_run(location = mtd_sample_data$spectraOrigin,
                            scan_polarity = ms_run_polarity)
    mtd <- rbind(mtd, mtd_msr)
    ## Add ms_runs[1-n] to correct sample in mtd_sample_data
    ms_runs_ref <- .mtd_get_field(mtd_msr, "^ms_run\\[\\d+\\]-location$",
                                exact = FALSE, fixed = FALSE)[[1]]
    mtd_sample_data$ms_runs_ref <- names(ms_runs_ref[match(ms_runs_ref,
                                                        mtd_sample_data$spectraOrigin)])
    mtd_sample_data$ms_runs_ref <- unlist(lapply(strsplit(mtd_sample_data$ms_runs_ref,
                                                    "-"), function(x){x[1]}))

    ## ASSAY
    ## sample_ref not mandatory
    mtd_a <- mtd_assay(assay = mtd_sample_data$sample_name,
                        sample_ref = mtd_sample_data$sample_ref,
                        ms_run_ref = mtd_sample_data$ms_runs_ref)
    mtd <- rbind(mtd, mtd_a)

    ## STUDY VARIABLES
    mtd_svar <- mtd_study_variables(mtd_sample_data,
                                    groups = groups,
                                    group_unit = group_unit,
                                    group_description = group_description,
                                    group_type = group_type,
                                    group_datatype = group_datatype,
                                    average_function = average_function,
                                    variation_function = variation_function,
                                    description = svar_description)
    mtd <- rbind(mtd, mtd_svar)

    mtd <- mtd_sort(mtd)
    mtd
}
```

##### MTD based on summarizedexperiment (Johannes code)

```{r}
library(SummarizedExperiment)

.mtd <- function(x, id = character(), software = character(),
                 studyVariables = colnames(colData(x)),
                 polarity = character()) {
    l <- ncol(x)
    sl <- seq_len(l)
    m <- mtd_skeleton(id = id, software = software)
    ## add samples; for now only the sample name.
    m <- rbind(m, mtd_sample(sample = rownames(colData(x))))
    ## add ms_run
    if (!length(polarity)) {
        ## guess polarity from sample name
        pos <- grepl("pos", rownames(colData(x)), ignore.case = TRUE)
        neg <- grepl("neg", rownames(colData(x)), ignore.case = TRUE)
        if (all(pos | neg)) {
            polarity <- rep("negative", l)
            polarity[which(pos)] <- "positive"
        }
    }
    m <- rbind(m, mtd_ms_run(location = rownames(colData(x)),
                             scan_polarity = polarity))
    ## add assay
    m <- rbind(m, mtd_assay(
                      assay = rownames(colData(x)),
                      sample_ref = paste0("sample[", sl, "]"),
                      ms_run_ref = paste0("ms_run[", sl, "]")))
    ## add study variables
    m <- rbind(m, mtd_study_variables(
                      data.frame(colData(x), check.names = FALSE),
                      groups = studyVariables))
    mtd_sort(m)
}
```

##### Add groupped optional variables

INSTRUMENT
```{r}
mtdInstrument <- function(x = matrix(), name = character(), source = character(),
                          analyzed = character(), detector = character(), add = TRUE) {
    if(!all(is.na(x))) {
        instr <- .mtd_get_field(x, "^instrument\\[\\d+\\]", exact = FALSE,
                                fixed = FALSE)[[1]]

        list_param <- c(list(name = name, source = source, detector = detector),
                        as.list(analyzed))

        if (!all(is.na(instr))) {
            if (add) {
                list_param <- Map(function(f) {
                                c(instr[grep(f, names(instr), fixed = TRUE)],
                                  list_param[[f]])},
                               names(list_param))
            }
            x <- x[!(x[, 1] %in% names(instr)), ]
        }
        new_instr <- do.call(mtd_fields, c(list(field_prefix = "instrument"),
                                            list_param))

        x <- mtd_sort(rbind(x, new_instr))
    }
    x
}
```

DATABASE
```{r}
mtdDatabase <- function(x = matrix(), name = character(), prefix = character(),
                        version = character(), uri = character(), add = TRUE) {
    if(!all(is.na(x))) {
        db <- .mtd_get_field(x, "^database\\[\\d+\\]", exact = FALSE,
                                fixed = FALSE)[[1]]

        list_param <- list(name = name, prefix = prefix,
                            version = version, uri = uri)

        if (!all(is.na(db))) {
            if (add & db[1] != "[,, \"no database\", null ]") {
                list_param <- Map(function(f) {
                                c(db[grep(f, names(db), fixed = TRUE)],
                                  list_param[[f]])},
                               names(list_param))
            }
            x <- x[!(x[, 1] %in% names(db)), ]
        }
        new_db <- .database(list_param$name, list_param$prefix,
                            list_param$version, list_param$uri)

        x <- mtd_sort(rbind(x, new_db))
    }
    x
}
```

CV
```{r}
mtdCv <- function(x = matrix(), label = character(), full_name = character(),
                  version = character(), uri = character(), add = TRUE) {
    if(!all(is.na(x))) {
        cv <- .mtd_get_field(x, "^cv\\[\\d+\\]", exact = FALSE,
                                fixed = FALSE)[[1]]

        list_param <- list(label = label, full_name = full_name,
                            version = version, uri = uri)

        if (!all(is.na(cv))) {
            if (add) {
                list_param <- Map(function(f) {
                                c(cv[grep(f, names(cv), fixed = TRUE)],
                                  list_param[[f]])},
                               names(list_param))
            }
            x <- x[!(x[, 1] %in% names(cv)), ]
        }
        new_cv <- .cv(list_param$label, list_param$full_name,
                        list_param$version, list_param$uri)

        ## do.call(mtd_fields, c(list(field_prefix = "cv"),
        ##                                              list_param))

        x <- mtd_sort(rbind(x, new_cv))
    }
    x
}
```

Contact
```{r}
mtdContact <- function(x = matrix(), name = character(),
                        affiliation = character(), email = character(),
                        add = TRUE) {
    if(!all(is.na(x))) {
        contact <- .mtd_get_field(x, "^contact\\[\\d+\\]", exact = FALSE,
                                    fixed = FALSE)[[1]]

        list_param <- list(name = name, affiliation = affiliation,
                            email = email)

        if (!all(is.na(contact))) {
            if (add) {
                list_param <- Map(function(f) {
                                c(contact[grep(f, names(contact), fixed = TRUE)],
                                  list_param[[f]])},
                               names(list_param))
            }
            x <- x[!(x[, 1] %in% names(contact)), ]
        }
        new_contact <- do.call(mtd_fields, c(list(field_prefix = "contact"),
                                            list_param))

        x <- mtd_sort(rbind(x, new_contact))
    }
    x
}
```

Single fields
```{r}
mtdField <- function(x = matrix(), field = character(), value = character(),
                        add = TRUE) {
    if(!all(is.na(x))) {
        if(!any(grepl(field, .MTD_FIELD_ORDER)))
            stop("Provide a valid MTD field.")

        existing_field <- .mtd_get_field(x, paste0("^", field, "\\[\\d+\\]"),
                                          exact = FALSE, fixed = FALSE)[[1]]

        if (!all(is.na(existing_field))) {
            if (add) {
                value <- c(existing_field, value)
            }
            x <- x[!(x[, 1] %in% names(existing_field)), ]
        }
        new_field <- mtd_fields(value, field_prefix = field)

        x <- mtd_sort(rbind(x, new_field))
    }
    x
}
```

#### Create SML

```{r}
sml_xcms_export <- function(xmse, mtd) {
    ## To implement since right now is not exported byt xcms
    abundance <- matrix("null", nrow = 1, ncol = nrow(sampleData(xmse)))
    sml <- sml_create(x = abundance)
    ## TODO: put a random reliability to pass validator.
    ## TODO: create issue in the valdiator
    sml$reliability <- 4
    sml <- sml_add_study_variable_columns(sml, mtd)
    sml[is.na(sml)] <- "null"
    sml
}

```

#### Create SMF

```{r}
matrix_integer_to_character <- function(m) {
    if (!is.matrix(m)) stop("Input must be a matrix")
    out <- matrix(as.character(m), nrow = nrow(m), ncol = ncol(m),
                    dimnames = dimnames(m))
    out[is.na(m)] <- "null"
    out
}

smf_xcms_export <- function(xmse) {
    ## use mz and rt information, don't consider the others
    feature_info <- featureDefinitions(xmse)
    ## abundance values:
    abundance <- matrix_integer_to_character(featureValues(xmse,
                                                           method = "sum"))

    charge = ifelse("charge" %in% names(feature_info), feature_info$charge,
                    character())
    adduct_ion = ifelse("adduct_ion" %in% names(feature_info),
                        feature_info$adduct_ion, character())
    isotopomer = ifelse("isotopomer" %in% names(feature_info),
                        feature_info$isotopomer, character())
    ## TODO: add SME_ID_REFS, SME_ID_REF_ambiguity_code
    smf <- smf_create(x = abundance,
                        exp_mass_to_charge = feature_info$mzmed,
                        retention_time_in_seconds = feature_info$rtmed,
                        retention_time_in_seconds_start = feature_info$rtmin,
                        retention_time_in_seconds_end = feature_info$rtmax,
                        charge = charge,
                        adduct_ion = adduct_ion,
                        isotopomer = isotopomer,
                        global_feature_id = rownames(feature_info))
    smf
}
```

#### Create SME

```{r}
sme_xcms_export <- function(xmse) {
    ## TODO: add SME. Default empty data.frame
    sme <- data.frame()
    sme
}
```

```{r}
updateMtdContent <- function(mtd, field, value) {
    idx <- which(mtd[, 1] == field)
    if(is.null(idx))
        stop("Field not detected. ...")

    mtd[idx, 2] = value
    mtd
}
```

##### Create unique table and save tsv

```{r}
## Generate the base MTD based on the xmse object with the mandatory fields
mtd <- .mtd(x = quantify(xmse),
            id = "test_jo_mtd",
            software = paste0("[MS, MS:1001582, xcms, ", packageVersion("xcms"), "]"),
            studyVariables = c("sample_group", "experiment", "sample_type"),
            polarity = "positive")
mtd <- updateMtdContent(mtd, "mzTab-ID", "xcms_mzTab-M_2-1_v2")

## Intrument
mtd <- mtdInstrument(x = mtd,
    name = "[MS, MS:1000449, LTQ Orbitrap,]",
    source = "[MS, MS:1000073, ESI,]",
    analyzed = c(`analyzer[1]` = "[MS, MS:1000291, linear ion trap,]"),
    detector = "[MS, MS:1000253, electron multiplier,]", add = TRUE)

## Database
mtd <- mtdDatabase(x = mtd,
                    name = "[MIRIAM, MIR:00100079, HMDB, ]",
                    prefix = "hmdb",
                    version = "3.6",
                    uri = "http://www.hmdb.ca/")

## CV
mtd <- mtdCv(x = mtd,
                label = "MS_test",
                full_name = "PSI-MS controlled vocabulary",
                version = "4.1.11",
                uri = "https://purl.obolibrary.org/obo/ms.obo")

## Contact
mtd <- mtdContact(x = mtd,
                    name = c("author_1","author_2"),
                    affiliation = c("EURAC","EURAC"),
                    email = c("author_1@mail.com","author_2@mail.com"))

## Single fields like "publication"
mtd <- mtdField(x = mtd, field = "publication",
                value = c("test pub","test pub 2"))

## Generate SML, SMF, SME
sml <- sml_xcms_export(xmse, mtd)
smf <- smf_xcms_export(xmse)
sme <- sme_xcms_export(xmse)

## Save file
path <- "./inst/mztabm/out"
if(!dir.exists(path))
    dir.create(path, showWarnings = FALSE)

filename = file.path(path,
                    paste0(.mtd_get_field(mtd, "^mzTab-ID$",
                                        exact = FALSE, fixed = FALSE),
                    ".mzTab"))

write.table(cbind(data.frame(rep("MTD", nrow(mtd))), mtd),
            file = filename, sep = "\t", row.names = FALSE,
            col.names = FALSE, quote = FALSE, fileEncoding = "UTF-8")
write.table("", file = filename, sep = "\t", row.names = FALSE,
            col.names = FALSE, append = TRUE, quote = FALSE,
            fileEncoding = "UTF-8")
suppressWarnings(write.table(sml, file = filename, sep = "\t",
                             row.names = FALSE, append = TRUE, quote = FALSE,
                             fileEncoding = "UTF-8"))
write.table("", file = filename, sep = "\t", row.names = FALSE,
            col.names = FALSE, append = TRUE, quote = FALSE,
            fileEncoding = "UTF-8")
if(nrow(smf)) {
    suppressWarnings(write.table(smf, file = filename, sep = "\t",
                                 row.names = FALSE, append = TRUE,
                                 quote = FALSE, fileEncoding = "UTF-8"))
    write.table("", file = filename, sep = "\t", row.names = FALSE,
                col.names = FALSE, append = TRUE, quote = FALSE,
                fileEncoding = "UTF-8")
}
if(nrow(sme)) {
    suppressWarnings(write.table(sme, file = filename, sep = "\t",
                                 row.names = FALSE, append = TRUE,
                                 quote = FALSE, fileEncoding = "UTF-8"))
}
```

### *Metabonaut* example file?

Here we would have some tentative annotations (only very few though) and a data
set with more phenotypic information (age, sex). Plus: we could add a dedicated
vignette with the mzTab-M export to *Metabonaut*.

# Importing through *RmzTabM*

# Session information

```{r}
sessionInfo()
```
