---
title: "Export the RIKEN_LIPIDOMICS test file to mzTab-M"
---

Kozo's example files used for the agentic export.

The data files are available
[here](https://github.com/kozo2/hupopsi-pymztabm-handson/tree/main/RIKEN_LIPIDOMICS).


# MTD + SML


Complications:

- we're missing a database identifier (for the annotations)
- we're missing an identifier or information on the reference annotation used;
  assuming it is an in-house database with standards.
- we don't have identifiers for the MS/MS spectra used for identification.

## MTD section

```{r}
library(RmzTabM)

#' Sample information
sd <- read.table("inst/tsv/1_Mouse_Adrenal_gland_1_forMTD.tsv", header = TRUE,
                 sep = "\t", quote = "", check.names = FALSE, row.names = 1L)
sd <- t(sd)

#' Abundance data and feature annotation
vals <- read.table("inst/tsv/1_Mouse_Adrenal_gland_1_forSMLSMF.tsv",
                   header = TRUE, sep = "\t", quote = "", check.names = FALSE)
all(rownames(sd) %in% colnames(vals))
fdef <- vals[, !colnames(vals) %in% rownames(sd)]
vals <- vals[, colnames(vals) %in% rownames(sd)]
```

For the present experiment we will encode MTD, SML and SME section. Let's start
with the general information in the MTD section. We use the the Schymanski score
for identification confidence level.

```{r}
mtd <- mtd_skeleton(
    id = "RIKEN_LIPIDOMICS",
    software = "[MS, MS:1003082, MS-DIAL, ]",
    database = "[,, in-house reference data, ]",
    small_molecule_identification_reliability =
        "[MS, MS:1002955, hr-ms compound identification confidence level,]")
mtd <- rbind(
    mtd,
    c("title", "Export of example file with RmzTabM"),
    c("description", "Example export of a RIKEN_LIPIDOMICS data set"),
    c("uri[1]", "https://metabography.riken.jp/menta.cgi/lipidomics/download_data_set"))
#' Manually add the BTO and NCBITaxon CVs
add_cv <- mtd_fields(
    label = c("BTO", "NCBITaxon"),
    full_name = c("The BRENDA Tissue Ontology (BTO)",
                  "NCBI organismal classification"),
    version = c("2021-10-26", "2025-12-03"),
    uri = c("https://www.ebi.ac.uk/ols4/ontologies/bto",
            "https://www.ebi.ac.uk/ols4/ontologies/ncbitaxon"),
    field_prefix = "cv")
add_cv[, 1L] <- sub("1", "4", add_cv[, 1L])
add_cv[, 1L] <- sub("2", "5", add_cv[, 1L])
mtd <- rbind(mtd, add_cv)
```

We next add sample information.

```{r}
mtd_s <- mtd_sample(
    sample = c("blank", "fads2_Adr_G_1", "fads2_Adr_G_2", "fads2_Adr_G_3",
               "WT_Adr_G_1", "WT_Adr_G_2", "WT_Adr_G_3"),
    species = c("null", rep("[NCBITaxon, NCBITaxon:10090, Mus musculus, ]", 6)),
    tissue = c("null", rep("[BTO, BTO:0000047, Adrenal gland, ]", 6))
)
mtd <- rbind(mtd, mtd_s)
```

Next we add the MS run information and assay information, assuming each sample
was measured in one run and hence each sample is assigned to one mzML file. We
could (should?) also add the MS runs mentioned in the `"Spectrum reference file
name"` column which contained the MS/MS used for annotation of the
compound. This would be needed if an SME section was provided, which we don't do
here (since we don't have a SMF section).

```{r}
msr <- union(rownames(sd), unique(fdef$`Spectrum reference file name`))

mtd_a <- rbind(
    mtd_ms_run(
        location = rownames(sd),
        format = "[MS, MS:1000584, mzML file, ]",
        id_format = "[MS, MS:1000530, mzML unique identifier, ]",
        scan_polarity = "positive"),
    mtd_assay(
        assay = rownames(sd),
        sample_ref = paste0("sample[", seq_len(nrow(sd)), "]"),
        ms_run_ref = paste0("ms_run[", seq_len(nrow(sd)), "]"))
)
mtd <- rbind(mtd, mtd_a)
```

Finally, we need to add the experimental information.

```{r}
mtd_svar <- mtd_study_variables(
    sd, groups = c("Genotype/Background", "Diet/Culture",
                   "Biological replicate")
)
mtd <- rbind(mtd, mtd_svar)
```

We have molecule annotations, so we must add also the `id_confidence_measure`
CV terms. Without much more information we guess that the data was matched
against on in-house library of standards. Thus RT + MS/MS matches would classify
the annotation as Schymanski level 1. Without RT or against only a reference
database it would be 2a. We have columns *Dot product* and *RT similarity*
available and we will use the CV terms `"[MS, MS:1003304, spectral dot product,
]"` and `"[,, RT similarity, ]"` for these.

```{r}
mtd <- rbind(
    mtd,
    mtd_fields("[MS, MS:1003304, spectral dot product, ]",
               "[,, RT similarity,]",
               field_prefix = "id_confidence_measure")
)
mtd <- mtd_sort(mtd)
```

OK, done with MTD.

## SML

For the SML we need also to provide the annotation/identification
confidence. This is encoded in column `"reliability"` for which we use the
Schymanski levels. As `id_confidence_measures` we have the MS/MS similarity and
the RT similarity (we use the first as the *best* confidence measure). Assuming
the annotation was performed against an in-house library of pure standards
(since RT are available and matched against) we assign a level 1 to all
annotations with RT and MS/MS matches.

Also, we add the following columns as `"opt_"` columns:

- `"Alignment ID"`
- `"INCHIKEY"`
- `"Average Mz"`
- `"Average Rt(min)"`
- `"Dot product"`
- `"RT similarity"`

```{r}
#' Defining the id reliability:
rel <- rep("5", nrow(fdef)) # m/z match
rel[fdef$`MS/MS matched`] <- "2a"
rel[fdef$`MS/MS matched` & fdef$`RT matched`] <- "1"

sml <- sml_create(
    x = vals,
    SMF_ID_REFS = rep("null", nrow(fdef)),        # no SMF section
    database_identifier = fdef$`Metabolite name`, # nothing else available
    chemical_formula = fdef$Formula,
    smiles = fdef$SMILES,
    chemical_name = fdef$`Metabolite name`,
    adduct_ions = fdef$`Adduct type`,
    reliability = rel,
    best_id_confidence_measure = "[MS, MS:1003304, spectral dot product, ]",
    best_id_confidence_value = fdef$`Dot product`,
    alignment_id = fdef$`Alignment ID`,
    inchikey = fdef$INCHIKEY,
    exp_mz = fdef$`Average Mz`,
    exp_rt = fdef$`Average Rt(min)`,
    similarity_ms2 = fdef$`Dot product`,
    similarity_rt = fdef$`RT similarity`)
#' Add study variable columns
sml <- sml_add_study_variable_columns(sml, mtd)
```

## Final export

```{r}
fl <- "inst/mztabm/out/riken-lipidomics-r-mtd-sml.mztab"
writeLines(c(paste("MTD", mtd[, 1L], mtd[, 2L], sep = "\t"), ""), con = fl)
suppressWarnings(
    write.table(sml, file = fl, append = TRUE,
                row.names = FALSE, quote = FALSE)
)
```


# MTD + SML + SMF + SME from `SummarizedExperiment`

Going the full path, putting the input data into a `SummarizedExperiment`.

```{r}
library(RmzTabM)

#' Sample information
sd <- read.table("inst/tsv/1_Mouse_Adrenal_gland_1_forMTD.tsv", header = TRUE,
                 sep = "\t", quote = "", check.names = FALSE, row.names = 1L)
sd <- t(sd)

#' Abundance data and feature annotation
vals <- read.table("inst/tsv/1_Mouse_Adrenal_gland_1_forSMLSMF.tsv",
                   header = TRUE, sep = "\t", quote = "", check.names = FALSE)
all(rownames(sd) %in% colnames(vals))
fdef <- vals[, !colnames(vals) %in% rownames(sd)]
vals <- vals[, colnames(vals) %in% rownames(sd)]

library(SummarizedExperiment)

#' rowData
rownames(fdef) <- fdef$`Alignment ID` # feature ID.

#' assay
vals <- as.matrix(vals)
rownames(vals) <- rownames(fdef)

se <- SummarizedExperiment(rowData = fdef, colData = sd, assay = list(vals))
```

## MTD

### Autogenerating MTD

```{r}
.mtd <- function(x, id = character(), software = character(),
                 studyVariables = colnames(colData(x)),
                 polarity = character()) {
    l <- ncol(x)
    sl <- seq_len(l)
    m <- mtd_skeleton(id = id, software = software)
    ## add samples; for now only the sample name.
    m <- rbind(m, mtd_sample(sample = rownames(colData(x))))
    ## add ms_run
    if (!length(polarity)) {
        ## guess polarity from sample name
        pos <- grepl("pos", rownames(colData(x)), ignore.case = TRUE)
        neg <- grepl("neg", rownames(colData(x)), ignore.case = TRUE)
        if (all(pos | neg)) {
            polarity <- rep("negative", l)
            polarity[which(pos)] <- "positive"
        }
    }
    m <- rbind(m, mtd_ms_run(location = rownames(colData(x)),
                             scan_polarity = polarity))
    ## add assay
    m <- rbind(m, mtd_assay(
                      assay = rownames(colData(x)),
                      sample_ref = paste0("sample[", sl, "]"),
                      ms_run_ref = paste0("ms_run[", sl, "]")))
    ## add study variables
    m <- rbind(m, mtd_study_variables(
                      data.frame(colData(x), check.names = FALSE),
                      groups = studyVariables))
    mtd_sort(m)
}

mtd <- .mtd(se, id = "my_study", software = "my software")
```

### Manual MTD completion

- need to provide annotation confidence and similar things.

```{r}
#' have a setter for each element/group?

mtd <- mtdInstrument(mtd, instrument, add = TRUE)
mtd <- mtdOntlogy()

#' or replacement methods?
ontology(mtd) # get the ontology part
ontology(mtd) <- "" # problem that we can't provide all parameters at once

mtdOntology(mtd, name = , uri = , ..., add = TRUE) # either adds it or replaces it
```

## SML

The problem is that there is no standard way in which the *annotation* is
encoded.

### Automatic generation

- take the `assay()` and encode that into the abundances.
- annotation *should* be provided in the `rowData`. But we need to configure
  which columns match which variables. Maybe the `translate` method could be
  handy here.

We add as much information to the `SummarizedExperiment`'s `rowData` to avoid
manual modification of the SML section as much as possible. In particular, we
add the *reliability* and the *best_id_confidence_measure* columns, the latter
as a CV param.

```{r}
#' Defining the id reliability:
rel <- rep("5", nrow(fdef)) # m/z match
rel[rowData(se)$`MS/MS matched`] <- "2a"
rel[rowData(se)$`MS/MS matched` & rowData(se)$`RT matched`] <- "1"

rowData(se)$reliability <- rel
rowData(se)$id_conf_measure <- "[MS, MS:1003304, spectral dot product, ]"
```

```{r}
#' @return a named `character` with names being the SML field names and values
#'     the *original* names e.g. the column names in a `data.frame` containing
#'     the respective information for the mzTab-M SML fields.
smlNameMapping <- function(SMF_ID_REFS = character(),
                           database_identifier = character(),
                           chemical_formula = character(), smiles = character(),
                           inchi = character(), chemical_name = character(),
                           uri = character(),
                           theoretical_neutral_mass = character(),
                           adduct_ions = character(), reliability = character(),
                           best_id_confidence_measure = character(),
                           best_id_confidence_value = character(), ...) {
    dots <- list(...)
    res <- c(SMF_ID_REFS = SMF_ID_REFS,
             database_identifier = database_identifier,
             chemical_formula = chemical_formula, smiles = smiles,
             inchi = inchi, chemical_name = chemical_name, uri = uri,
             theoretical_neutral_mass = theoretical_neutral_mass,
             adduct_ions = adduct_ions, reliability = reliability,
             best_id_confidence_measure = best_id_confidence_measure,
             best_id_confidence_value = best_id_confidence_value)
    if (length(names(dots))) {
        dots <- dots[names(dots) != ""]
        res <- c(res, unlist(dots, recursive = FALSE))
    }
    res
}

#' Get an SML section from a `SummarizedExperiment` `x`.
#'
#' @param x `SummarizedExperiment` with `rowData()` containing annotation
#'     information and `assay()` the abundance values.
#'
#' @param smlNameMapping named `character()` defining which columns in
#'     `rowData()` are used for which annotation field in the SML table. See
#'     details for more information.
#'
#' @return SML `data.frame`
#'
#' @noRd
.sml_from_se <- function(x, smlNameMapping = smlNameMapping()) {
    vals <- assay(x)
    rd <- data.frame(rowData(x), check.names = FALSE)
    if (any(miss <- !smlNameMapping %in% colnames(rd)))
        stop("Column(s) ",
             paste0("\"", smlNameMapping[miss], "\"", collapse = ", "),
             " provided with 'smlNameMapping' not found in 'rowData(x)'.",
             call. = FALSE)
    rl <- as.list(rd[, smlNameMapping, drop = FALSE])
    names(rl) <- names(smlNameMapping)
    do.call("sml_create", c(list(x = assay(x)), rl))
}

smlMap <- smlNameMapping(
    chemical_formula = "Formula", smiles = "SMILES",
    chemical_name = "Metabolite name", adduct_ions = "Adduct type",
    reliability = "reliability", best_id_confidence_value = "Dot product",
    best_id_confidence_measure = "id_conf_measure",
    alignment_id = "Alignment ID")

sml <- .sml_from_se(se, smlMap)
sml <- sml_add_study_variable_columns(sml, mtd)

```

### Manual completion

Need to also cross-reference MTD.


## SMF

### Automatic generation

```{r}
smfNameMapping <- function(exp_mass_to_charge = character(),
                           retention_time_in_seconds = character(),
                           retention_time_in_seconds_start = character(),
                           retention_time_in_seconds_end = character(),
                           SME_ID_REFS = character(),
                           SME_ID_REF_ambiguity_code = character(),
                           charge = character(), adduct_ion = character(),
                           isotopomer = character(), ...) {
    dots <- list(...)
    res <- c(exp_mass_to_charge = exp_mass_to_charge,
             retention_time_in_seconds = retention_time_in_seconds,
             retention_time_in_seconds_start = retention_time_in_seconds_start,
             retention_time_in_seconds_end = retention_time_in_seconds_end,
             SME_ID_REFS = SME_ID_REFS,
             SME_ID_REF_ambiguity_code = SME_ID_REF_ambiguity_code,
             charge = charge, adduct_ion = adduct_ion,
             isotopomer = isotopomer)
    if (length(names(dots))) {
        dots <- dots[names(dots) != ""]
        res <- c(res, unlist(dots, recursive = FALSE))
    }
    res
}

#' Maybe have a `defaultSmfNameMapping("xcms")` that provides some fields.

.smf_from_se <- function(x, smfNameMapping = smfNameMapping()) {
    vals <- assay(x)
    rd <- data.frame(rowData(x), check.names = FALSE)
    if (any(miss <- !smfNameMapping %in% colnames(rd)))
        stop("Column(s) ",
             paste0("\"", smfNameMapping[miss], "\"", collapse = ", "),
             " provided with 'smfNameMapping' not found in 'rowData(x)'.",
             call. = FALSE)
    rl <- as.list(rd[, smfNameMapping, drop = FALSE])
    names(rl) <- names(smfNameMapping)
    do.call("smf_create", c(list(x = assay(x)), rl))
}

rowData(se)$rtmed <- rowData(se)$`Average Rt(min)` * 60

smfMap <- smfNameMapping(exp_mass_to_charge = "Average Mz",
                         retention_time_in_seconds = "rtmed",
                         adduct_ion = "Adduct type",
                         alignment_id = "Alignment ID")

smf <- .smf_from_se(se, smfMap)
```

### Manual completion

Should be all done.


## SME

### Automatic generation

### Manual completion

# Session information

```{r}
sessionInfo()
```
